Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2016-05-10 pkg/glmertree/paper/treatment_interactions.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@848 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nobf,doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{ctable}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{lineno}
\usepackage{afterpage}
\linespread{1.2}
\usepackage{tabto}

@@

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. Several tree-based algorithms for the detection of treatment-subgroup interactions have been developed (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}, see \citeNP{DoovyDuss14} for an overview). Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible method for estimating so-called optimal treatment regimes, which allows users to select from a range of statistical methods, including trees, to derive the optimal regimes. As such, the \citeA{ZhanyTsiaStat} method also provides a tree-based method for treatment-subgroup interaction detection.

@@

\begin{table}[]
	\caption{Node specific summary statistics and effect sizes of HAM-D scores for GLM tree}
	\begin{tabular}{lcccc}
		\thickline
		Node	&	$\hat{\mu}_{CBT}$	& $\hat{\mu}_{PHA}$	& pooled SD	& Cohen's $d$	\\
		\hline
		2		&	13.25		& 10.29		& 7.52		& 0.39			\\
		4		&	7.82		& 7.55		& 5.51		& 0.05			\\
		5		&	10.35		& 7.69		& 6.64		& 0.40			\\
		\hline
		\multicolumn{5}{l}{\small \textit{Note.} $\hat{\mu}$ values are predicted posttreatment HAM-D}\\
		\multicolumn{5}{l}{\small means, evaluated at the global HAM-D pretreatment}\\
		\multicolumn{5}{l}{\small mean. Pooled SDs are calculated based on post-}\\
		\multicolumn{5}{l}{\small treatment HAM-D scores.}\\ 
	\end{tabular}
	\label{tab:lmtree_C&W}
\end{table}

\begin{table}[]
	\caption{Node specific summary statistics and effect size for GLMM tree}
	\label{tab:lmertree_C&W}
	\begin{tabular}{lcccc}
	\thickline
		Node	&	$\hat{\mu}_{CBT}$	& $\hat{\mu}_{PHA}$	& pooled SD	& Cohen's $d$	\\
		\hline
		2		&	8.27		& 7.99		& 5.86		& 0.05			\\
		3		&	10.84		& 8.21		& 6.80		& 0.39			\\
		\hline
		\multicolumn{5}{l}{\small \textit{Note.} $\hat{\mu}$ values are predicted posttreatment HAM-D}\\
		\multicolumn{5}{l}{\small means, evaluated at the global HAM-D pretreatment}\\
		\multicolumn{5}{l}{\small mean. Pooled SDs are calculated based on post-}\\
		\multicolumn{5}{l}{\small treatment HAM-D scores.}\\ 
	\end{tabular}
\end{table}

@@

\section{Discussion}

@@

These findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures. However, it should be noted that the simulations show that GLMM tree performs very well, if the model is correctly specified. That is, if there are subgroups with respect to the partitioning variables, so that there are different parameters of the GLM in each of these subgroups, then GLMM tree will accurately recover those subgroups. However, misspecification of the model can reduce performance. One source of misspecification would be the omission of relevant variables in the GLM, or as potential partitioning variables. For example, if there are actual subgroups, but the variables describing them are not entered as partitioning variables, the algorithm can only approximate the subgroups using variables that are available for partitioning. Another source of misspecification would be the inclusion of irrelevant variables, either in the linear predictor of the GLM or as partitioning variables, which may reduce the power to detect the actual subgroups. It should be noted, however, that the number of partitioning variables did not substantially influence performance of the algorithm(s) in our simulations.

>>>>

\documentclass[nobf,doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{ctable}
\usepackage{rotating}
\usepackage{lineno}
\usepackage{placeins}
\linespread{1.2}
\usepackage{tabto}

@@

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. Several tree-based algorithms for the detection of treatment-subgroup interactions have been developed (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}, see \citeNP{DoovyDuss14} for an overview). Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible method for estimating so-called optimal treatment regimes, which allows users to select from a range of statistical methods, including trees, to derive the optimal regimes. As such, the \citeA{ZhanyTsia12Stat} method also provides a tree-based method for treatment-subgroup interaction detection.

@@

The GLM tree (Figure~\ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education. Node 2 of Figure~\ref{fig:lmtree_C&W} indicates that for patients with a low level of education, antidepressant medication provides the greatest reduction in HAM-D scores. Table~\ref{tab:lmtree_C&W} shows that the treatment effect difference in Node 2 amounts to a small to medium effect size (Cohen's $d = 0.39$). Node 4 indicates that for patients with a higher level of education, and no comorbid anxiety disorder, the reduction in HAM-D scores is about the same for CBT and PHA. Node 5 indicates, that for patients with a higher level of education and a comorbid anxiety disorder, the reduction in HAM-D scores is greatest for PHA. In Node 5, a small to medium effect size was also observed (Cohen's $d = 0.40$; Table~\ref{tab:lmtree_C&W}).

@@

\clearpage

\begin{table}[!b]
	\caption{Node specific summary statistics and effect sizes of HAM-D scores for GLM tree}
	\begin{tabular}{lcccc}
		\thickline
		Node	&	$\hat{\mu}_{CBT}$	& $\hat{\mu}_{PHA}$	& pooled SD	& Cohen's $d$	\\
		\hline
		2		&	13.25		& 10.29		& 7.52		& 0.39			\\
		4		&	7.82		& 7.55		& 5.51		& 0.05			\\
		5		&	10.35		& 7.69		& 6.64		& 0.40			\\
		\hline
		\multicolumn{5}{l}{\small \textit{Note.} $\hat{\mu}$ values are predicted posttreatment HAM-D}\\
		\multicolumn{5}{l}{\small means, evaluated at the global HAM-D pretreatment}\\
		\multicolumn{5}{l}{\small mean. Pooled SDs are calculated based on post-}\\
		\multicolumn{5}{l}{\small treatment HAM-D scores.}\\ 
	\end{tabular}
	\label{tab:lmtree_C&W}
\end{table}

\begin{table}[!b]
	\caption{Node specific summary statistics and effect size for GLMM tree}
	\label{tab:lmertree_C&W}
	\begin{tabular}{lcccc}
		\thickline
		Node	&	$\hat{\mu}_{CBT}$	& $\hat{\mu}_{PHA}$	& pooled SD	& Cohen's $d$	\\
		\hline
		2		&	8.27		& 7.99		& 5.86		& 0.05			\\
		3		&	10.84		& 8.21		& 6.80		& 0.39			\\
		\hline
		\multicolumn{5}{l}{\small \textit{Note.} $\hat{\mu}$ values are predicted posttreatment HAM-D}\\
		\multicolumn{5}{l}{\small means, evaluated at the global HAM-D pretreatment}\\
		\multicolumn{5}{l}{\small mean. Pooled SDs are calculated based on post-}\\
		\multicolumn{5}{l}{\small treatment HAM-D scores.}\\ 
	\end{tabular}
\end{table}

\FloatBarrier
\section{Discussion}

@@

These findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures. However, it should be noted that the simulations show that GLMM tree performs very well, if the model is correctly specified. That is, if there are subgroups with respect to the partitioning variables, so that there are different parameters of the GLM in each of these subgroups, then GLMM tree will accurately recover those subgroups. However, misspecification of the model can reduce performance. One source of misspecification would be the omission of relevant variables in the GLM, or as potential partitioning variables. For example, if there are actual subgroups, but the variables describing them are not entered as partitioning variables, the algorithm can only approximate the subgroups using variables that are available for partitioning. Another source of misspecification would be the inclusion of irrelevant variables, either in the linear predictor of the GLM or as partitioning variables, which may reduce the power to detect the actual subgroups. However, the number of partitioning variables did not substantially influence performance of the algorithm(s) in our simulations.

As discussed in the Introduction, treatment-subgroup interactions have gained interest from several areas, and several different tree-based methods for their detection have been developed. In the current study, we have limited our comparison to the performance of GLM and GLMM-based MOB. Obviously, further research on what the different methods for treatment-subgroup interaction detection have in common and what sets them apart is needed.
